---
title: "DATA607_Assignmnet3 "
author: "Jerald Melukkaran"
date: "2025-02-16"
output: html_document
---
# Overview 


### Initial setup

Here we load all the necessary libraries 

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readr)

```

### Loading data into dataframe 

Obtained the permalink for the file from github and loaded the csv into a dataframe 

```{r message=FALSE, warning=FALSE}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/majors-list.csv"
majors <- read_csv(url)

```

#### Q1 
Finding majors that have DATA or STATISTICS in them 

```{r message=FALSE, warning=FALSE}
q1majors <- majors %>% 
  filter(grepl("DATA|STATISTICS", Major, ignore.case = TRUE))

print (q1majors)

```

#### Q2 

```{r}
fruits <- c("bell pepper", "bilberry", "blackberry", "blood orange",
            "blueberry", "cantaloupe", "chili pepper", "cloudberry",
            "elderberry", "lime", "lychee", "mulberry",
            "olive", "salal berry")

cat("c(", paste(shQuote(fruits, type="cmd"), collapse = ", "), ")")
```

We start with the c( , shQuote adds quotes, cmd makes it double quotes, collapse separates the elements with a comma and a space and we end with the ) 


#### Q3


1. (.)\1\1    
Matches characters that are consecutively repeated thrice.  
eg aaa, hhh

2. (.)(.)\\2\\1  
Gets two characters, and checks if the two characters repeat in reverse.   
eg xyyx

2. (..)\1
Gets two characters and checks if they repeat again without any other characters in between   
eg xyxy , k5k5


3. (.).\\1.\\1
Matches a 5 character sequence where the 1st,3rd and 5th characters are the same. 2nd and 4th can be any character  
eg xyx8x , 4d424

4. (.)(.)(.).*\\3\\2\\1  
Gets sequences that start with 3 characters and end with the same three characters in reverse with any number of characters in between.  
eg abcdefhcba , 14i29dhfi41



#### Q4 

1. Starting and ending with same character  
ANS: "(.).*\1$"

2. Words that contain repeated pairs  
ANS: "(..).*\1" 

3. Contains 1 letter repeated in 3 places   
ANS "(.).* \1.*\1" 









