---
title: "Week3GlobalBaselineEstimate "
author: "Jerald Melukkaran"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(readr)
library(tidyverse)
```

## Introduction 
For this exercise im trying to use global baseline estimates to help us predict movie ratings. 

We first start with the user reviews, initially a long list of the ratings that were given to the movies from 1-5 converted to a wider format using pivot_wider for better viewing 

```{r echo=TRUE, message=FALSE, warning=FALSE}
viewer_ratings <- read_csv ("viewer_ratings.csv")

ratings <- viewer_ratings %>%
  pivot_wider(
    names_from = "movie_name",
    values_from = "viewer_rating",
  )

knitr::kable((ratings), "simple")
```

Now we add averages for each movie in a column, and viewer averages for each row

```{r echo=TRUE, message=FALSE, warning=FALSE}
ratings$viewer_avg <- round(rowMeans(ratings[, -1], na.rm = TRUE),1)
movie_avg <- round(colMeans(ratings[, -1], na.rm = TRUE), 1)
ratings <- rbind(ratings, c("movie_avg", movie_avg))

knitr::kable((ratings), "simple")
```


## Global estimate 

THe formula for predicting what someones rating for a specific movie according to the the spreadsheet was 
Global Baseline Estimate = Mean Movie Rating + (The specific movies avg rating - MMR) + (The specific viewers average rating - MMR) 

Now we can predict how the viewers who did not see the movie might rate it. And replace the NA values with the GLobal estimates. 

```{r}

MMR <- 3.7
ratings_data <-ratings


for (movie in names(ratings_data)[2:7]) {  # Loop through columns 2 - 7 
  for (i in 1:nrow(ratings_data)) {  # Loop through rows 
    if (is.na(ratings_data[i, movie])) {
      # Get movie's average rating 
      movie_avg_rating <- as.numeric(ratings_data[nrow(ratings_data), movie])
      # Get viewers average rating
      viewer_avg_rating <- as.numeric(ratings_data[i, 8])
      # Calculate Global Baseline EstimateS
      baseline_estimate <- MMR + (movie_avg_rating - MMR) + (viewer_avg_rating - MMR)
      # Replace the NA value with estimates
      ratings_data[i, movie] <-as.character(baseline_estimate) 
    }
  }
}

knitr::kable((ratings_data), "simple")


```




