---
title: "DATA607_Project1"
author: "Jerald Melukkaran"
date: "23/2/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
```

## Introduction 

In this project we clean up a particularly formatted text document for a chess tournament and extract the average opponent pre-rating for each player. 



#### Import and cleanup

Steps   

1) We import the textfile from github where its been stored, remove the dotted lines, and then combine the two rows per player into 1 row 

```{r echo=TRUE, warning=FALSE}
raw_data <- readLines("https://raw.githubusercontent.com/jerryjerald27/Data-607/refs/heads/main/Week4Assignment/tournamentinfo.txt")[-(1:3)]
#remove the dotted lines 
raw_data <- raw_data[str_detect(raw_data, '^\\-+$') == FALSE]
# #combine two lines to one row 
combined_data <- raw_data %>%
  str_trim() %>%
  # .[str_detect(., "^\\-S")] %>%  
  enframe(name = NULL) %>%
  mutate(row_num = rep(1:(n()/2), each = 2)) %>%
  group_by(row_num) %>%
  summarise(combined = paste(value, collapse = " ")) %>%
  pull(combined)

knitr::kable((head(combined_data)),"simple")
```

2) We then map this combined long observation into each of their specific columns 

```{r}
# Split combined data by "|" and create a dataframe
raw_data_frame <- combined_data %>%
  str_split(pattern = "\\|") %>%
  map_dfr(~tibble(Player = .[1], Name = .[2], State =.[11], Total = .[3], Pre_Rating = .[12], 
                  Round1 = .[4], Round2 = .[5], Round3 = .[6], 
                  Round4 = .[7], Round5 = .[8], Round6 = .[9], Round7 = .[10]))

knitr::kable(head(raw_data_frame),"simple")

```

3) We can now mutate the columns to further clean up the Pre rating column to only include the rating, and the Rounds column to only include the Opponent ID and not the game result 

```{r}
raw_data_frame <- raw_data_frame %>%
  mutate(
    Pre_Rating = as.numeric(str_extract(Pre_Rating, "(?<=R:\\s{1,2})\\d{3,4}")),   # Gets just the prerating from the prerating column
    across(starts_with("Round"), ~ gsub("[^0-9]", "", .)),    #mutates all columns starting with rounds to only include digits
    Player = as.numeric(Player)
  )

knitr::kable(head(raw_data_frame),"simple")



```

#### Calculating average Opponent ratings 

This is done in 2 steps  

1) For each player we replace the opponent Ids for each of the rounds with the Pre_ratings of the opponents 


```{r}
round_columns <- paste0("Round", 1:7)

for (col in round_columns) {
  raw_data_frame[[col]] <- sapply(raw_data_frame[[col]], function(x) {
    rating <- raw_data_frame$Pre_Rating[raw_data_frame$Player == x]
    if (length(rating) > 0) rating else NA
  })
}

knitr::kable(head(raw_data_frame),"simple")
```

2) We then Average out the ratings across the rounds
```{r}
# Calculate average opponent pre-rating
raw_data_frame$Avg_Opp_Pre_rating <- round(rowMeans(raw_data_frame[round_columns], na.rm = TRUE))
Average_chess_rating <- raw_data_frame %>% select(Name, State, Total, Pre_Rating, Avg_Opp_Pre_rating)
 write.csv(Average_chess_rating, "cleaned_chess.csv", row.names = TRUE)
```


### Conclusion 

We can now view the completed table and export it into a .csv file 

```{r}
knitr::kable((Average_chess_rating), "simple")